\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-05-03 trunk/europa.tex
updates

\eu is implemented as a library that encapsulates all elements
described above making them accessible via modeling and Application
Programming Interface (API) layers.  A user will typically use the
modeling facilities to inject a problem domain description and the
API to extract and manipulate plans and to obtain information about
plan flaws and constraint violations at any stage of the planning
process.

\begin{figure}[b]
\centering
\includegraphics[scale=0.4]{figs/EUROPA-Architecture.pdf}
\caption{\small The \eu software architecture.}
\label{fig:europa-architecture}
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships. The \textbf{Constraint
  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
constraints that define relationships among them. It also provides an
efficient arc consistency mechanism \cite{mackworth77}. The
\texttt{CRE} is designed so that specialized reasoning algorithms for
new constraints can be easily and efficiently plugged in. The
\textbf{Plan Database} (\texttt{PDB}) manages object and token
types. It also maintains the state of partial plans in terms of
object, variable and token instances. This constitutes the data
foundation on which end user applications and automated problem solvers
 are built. The \textbf{Rules Engine} (\texttt{RE})
modularly adds the semantics of domain rules that describe
dependencies between tokens. The \textbf{Solver Module} provides the
framework for flaw detection and resolution that is needed to
implement search algorithms for planning. The \textbf{API Layer}
provides access to services from all the different modules in a
programmatic way so that \eu can be easily embedded into specialized
applications.

Describing problem domains and problem instances in terms of objects,
variables and domain rules in a programmatic way is cumbersome to the
extent of being impractical; this motivated the creation of modeling
languages that support those definitions in a much more natural and
compact way. \texttt{NDDL} (New Domain Description Language
[pronounced ``noodle'']) \cite{NDDL} is the main modeling language for
\eu and will be described in detail in the Modeling section
below. Another, \texttt{ANML} is a language under development that
provides a more intuitive syntax \cite{smith08}.

@@

\item \textit{Temporal Network}: Extensions to the \texttt{CRE} module to
  reason about temporal constraints

\item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
  and Solver modules to provide reasoning and search for metric
  resources.

\item \textit{Chronological Backtracking Solver}: A general purpose
  built-in solver to provide the ability to plan out-of-the-box.
  Problems that require challenging search may require specialized
  solvers to be built as explained in Section \ref{sec:europa:search}.

@@

@@

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different domains and problems, simply by providing
different models and goals. Consequently the expressiveness of the
language for models, goals and plans is of great importance. For \eue,
\texttt{NDDL} allows the user to specify the different components of a
problem domain in a precise and concise manner. Among the main
features that make \texttt{NDDL} a powerful tool for describing
problem domains are that it is object oriented. \texttt{NDDL} supports
classes and inheritance in similar fashion to popular Object-Oriented
languages like Java and C++. It also supports polymorphism for some of
the planning components. Using object classes and instances is a
time-tested approach to naturally describe problem domains
\cite{rumbaugh}. It offers declarative constructs to define
constraints and causality dependencies between actions and effects and
finally it offers procedural constructs to populate \eus \texttt{PDB}
with specific problem instances expressed in terms of objects,
variables, constraints, facts and goals. To articulate the most important
elements of \texttt{NDDL}, we show how the Shopping Agent problem
could be modeled:

@@

\item[\textbf{Object Types}:] \texttt{NDDL} supports an object-oriented
  approach, where common data and behavior can be abstracted as a
  class. In this case, we have the \textit{Product} class represent
  the products that the Shopping Agent needs, \textit{Location}
  represents the places where products can be found,
  \textit{ProductLocation} represents the association arising when a
  product can be purchased at a location. The Shopping Agent itself is
  represented by an \textit{Agent} class, which relies on the
  \textit{AgentLocation} class to keep track of where the agent is at
  any point in time. \texttt{NDDL} supports a single rooted class
  hierarchy so that state and behavior can be extended through
  inheritance. We take advantage of that feature by making
  \textit{AgentLocation} inherit from \eus built-in \textit{Timeline}
  class which doesn't allow predicates to overlap in time for the same
  object instance. This feature is used for instance, to ensure that
  an agent can only be at one place at a time in our example.

@@

Domain rules can be defined in \texttt{NDDL} in the context of a
predicate or an action. In the example above we can see the main
elements of a rule definition:

@@

An \eu user can introduce new constraint types by implementing
specialized constraint and propagator classes; however, \eu provides a
built-in constraint library with a range of arithmetic, temporal, set
and resource constraint types so that a large number of problem
domains can be modeled without having to introduce new constraint
types. Tables \ref{tab:calconst}, \ref{tab:setconst},
\ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
types included in \eus Constraint Library.

Temporal and resource constraints are especially important to model
real-world planning domains.

\eu uses variables to explicitly represent timepoints for plan
actions and states. Constraints among timepoints provide a natural
way to express domain axioms. For example, in order to state that
activity $A$ must occur before activity $B$ we can say that the end
timepoint of $A$ is $leq$ the start timepoint of $B$.
\cite{dechter91} proposed that constraints among timepoints can be
grouped together to form a Simple Temporal Network
(\texttt{STN}). Such a network can be transformed into a Distance
Graph (\texttt{DG}) where the outward arc from a node represents the
maximum distance from the source node to the target node.
Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
a single constraint along with the resulting \texttt{DG}.
\cite{dechter91} also showed that shortest path algorithms could be
used to propagate values in the network and discover a negative cycle:
namely a path from a node to itself that has a path length less than
$0$. If such a cycle exists, the network is inconsistent. It was
further shown that a single-source shortest path algorithm was
sufficient to detect a negative cycle and provide sufficient
propagation to yield a backtrack-free search. \eus propagator for
temporal constraints is a straight implementation of this algorithm.

\begin{figure}[!htb]
  \centering
  \subfloat[\small A Simple Temporal Network (STN)]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
  \subfloat[\small Resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
  \caption{\small An illustration of a STN and its corresponding
    distance graph.}
  \label{fig:stn}
\end{figure}

Most real-world planning problems need to deal with plans made up of
actions that typically depend upon limited resources (\eg fuel,
storage capacity, battery state of charge, data buffers).  \eu
provides built-in support for resource computation. In \eus
representation resource object types come in three forms with related
constraints. \textbf{Reservoir} resources can have both
\texttt{consume(resource,time,quantity)} and
\texttt{produce(resource,time,quantity)} constraints; production and
consumption is instantaneous. \textbf{Reusable} resources on the other
hand, have constraints that use a quantity of the resource for a
specific duration (\texttt{uses(resource,start,duration,quantity)}),
which are then subsequently released. \textbf {Unary} resources are
reusables with capacity$ = 1$, which allows \eu to perform run time
savings in comparison to the generic reusable class. Configuration
options are offered such that in addition to the usual constraint
violations when a resource is completely depleted, plan flaws can be
generated whenever a resource level is outside a specific upper or
lower bound, or when instantaneous production or consumption exceeds a
specific threshold.

Since \eu supports a flexible temporal representation, where time is
represented by intervals, instead of single values, determining
whether a resource is the source of a plan flaw or a constraint
violation in general is substantially more difficult than just
computing a profile of the resource and checking against some limits.
With temporal flexibility, a number of resource profiles could be
possible depending on precisely when each production/consumption event
occurs out of a range of possible values that are in the domain of the
variables that represent the temporal extent for the event.
\cite{Muscettola04,Muscettola06} proposed an algorithm where the
variables that represent the time for production/consumption events
are placed in a temporal network, similar to the network used by the
temporal propagator; constraints on the resource can then be seen as
quantities flowing through that network. Computing the maximum flow
then, yields the maximum resource usage and thus a lower bound on the
resource level; computing a minimum flow will similarly yield an upper
bound. Since maximum flow on a temporal network can be computed in
polynomial time, \eu takes advantage of this algorithm to efficiently
compute upper and lower bound envelopes on the resource
level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
computation.

\begin{figure}
\centering
\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
\caption{\small An example of resource envelopes computed by \eu.}
\label{fig:resenvelopes}
\end{figure}

Resource envelopes can be useful when trying to evaluate plan
brittleness in the face of changing activity durations or start
times. They can also be useful when searching for plans; however they
are relatively expensive to compute. Therefore their ability to prune
the search space must be weighted against the ability to search faster
\cite{6007772, morris11}.

@@

Open conditions and threats allow flaw detection and resolution at a
higher-level of abstraction (\ie in terms of objects and tokens) than
that of binding variables as is common in Constraint Satisfaction
Problems. This is advantageous when a solver applies heuristics for
ordering choices since it provides a richer context in which to make
decisions.  Furthermore, it aids in reducing the amount of work done
by a solver so that only the necessary refinements are made leaving a
partial plan with sufficient flexibility. For example, one can bypass
committing unbound timepoint variables since threats will force a
solver to impose ordering restrictions on these variables.  Thus the
planning process may yield partially-ordered plans for which all
possible extensions are provably valid.

The Plan Space Planning algorithm described above can be implemented
in multiple ways depending on the approach chosen for flaw and
resolver selection and for backtracking. \eus built-in solver
implements a chronological backtracking algorithm that is summarized
in Algorithm \ref{alg:europa:solve}.

\begin{algorithm}[H]
\KwIn{PartialPlan $plan$}
\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
\Begin{
  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
  \BlankLine
  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
  \BlankLine
  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \; \label{li:decisionpt}
  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
    \lElse{ $decision.undo()$ \; } \label{li:backtrack}
   }
   \Return \texttt{false} \; \label{li:noplan}
 }
\caption{$\mathrm{bool} ~ solve(plan)$}
\label{alg:europa:solve}
\end{algorithm}

@@

>>>>

\eu is implemented as a library that encapsulates all elements
\kcomment{previously described} and available via modeling and
Application Programming Interfaces (APIs).  A user will typically use
modeling facilities to inject a problem domain description and the API
to extract and manipulate plans and to obtain information about plan
flaws and constraint violations at any stage of the planning process.

\begin{figure}[b]
\centering
\includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
\caption{\small The \eu software architecture.}
\label{fig:europa-architecture}
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships. The \textbf{Constraint
  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
constraints that define relationships among them. It also provides an
efficient arc consistency mechanism \cite{mackworth77} designed so
that specialized reasoning algorithms for new constraints can be
easily and efficiently plugged in. The \textbf{Plan Database}
(\texttt{PDB}) manages object and token types. It also maintains the
state of partial plans in terms of object, variable and token
instances. This constitutes the data foundation on which end user
applications and automated problem solvers are built. The
\textbf{Rules Engine} (\texttt{RE}) adds the semantics of domain rules
that describe dependencies between tokens. The \textbf{Solver Module}
provides the framework for flaw detection and resolution that is
needed to implement search algorithms for planning. The \textbf{API
  Layer} provides access to services from all the different modules in
a programmatic way so that \eu can be easily embedded into specialized
applications.

Describing problem domains and problem instances in terms of objects,
variables and domain rules is cumbersome to the extent of being
impractical; this motivated the creation of modeling languages that
support those definitions in a much more natural and compact way. \nd
(New Domain Description Language [pronounced ``noodle'']) \cite{NDDL}
is the \kcomment{primary} modeling language for \eu described in
Section \ref{sec:europa:modeling}. Another, \texttt{ANML}
\cite{smith08} is a language under development that provides a more
intuitive syntax \kcomment{and out of scope for this chapter}.

@@

\item \textit{Temporal Network}: Extensions to the \texttt{CRE} module to
  reason about temporal constraints.

\item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
  and \textbf{Solver Module} to provide reasoning and search for
  metric resources.

\item \textit{Chronological Backtracking Solver}: A general purpose
  built-in solver to provide the ability to plan out-of-the-box.
  Problems that require challenging search may require specialized
  solvers to be built as noted in Section \ref{sec:europa:search}.

@@

\kcomment{Before going into the details of each module, a simple
  example illustrates how they interact at an abstract level:}

class LightBulb extends Timeline
{
    predicate On  {}
    predicate Off {}
}

class LightSwitch extends Timeline
{
    LightBulb myBulb_;

    LightSwitch(LightBulb b)
    {
        myBulb_ = b;
    }

    action turnOn  { duration=1; }
    action turnOff { duration=1; }
}

LightSwitch::turnOn
{
    met_by(condition object.myBulb_.Off);  // Bulb must be Off to be turned On
    meets(effect object.myBulb_.On);   // Must be turned on through the switch
}

LightSwitch::turnOff
{
    met_by(condition object.myBulb_.On);   // Bulb must be On to be turned Off
    meets(effect object.myBulb_.Off);      // Must be turned off through the switch
}

// Problem instance : turning the light off
LightBulb bulb1 = new LightBulb();
LightSwitch switch1 = new LightSwitch(bulb1);

// At time 0, the bulb is on
fact(bulb1.On initialCondition);
eq(initialCondition.start,0);

// We want the bulb to be off by time 10
goal(bulb1.Off goal1);
lt(0,goal1.start);
lt(goal1.start,10);

\kcomment{The example above is a \nd (covered in more detail in
  Section \ref{sec:europa:modeling}) model of a light bulb and a
  switch that controls it along with a problem instance that requires
  the light bulb to be \texttt{off} by time$=10$, given an initial
  state with the bulb \texttt{on} at time$=0$. The \nd module allows
  \eu to ingest this model and the problem description. Once the \nd
  statements are parsed, different object types, object instances and
  constraints are instantiated inside the \texttt{PDB} and
  \texttt{CRE} modules where they are available for the user and the
  solver modules to query and manipulate programmatically. This
  instantiation of the initial state constitutes the initial plan that
  \eu holds for this particular problem instance.

  The user can then invoke functionality available in the solver
  module to modify or extend this initial plan in a way that
  eliminates any flaws (violated constraints, unsatisfied goals,
  oversubscribed resources). When actions are added to a plan, the
  \texttt{RE} ensures that domain rules expressed in the model are
  enforced when \eu creates new action or predicate instances. For
  instance, if a \texttt{LightSwitch::turnOn} action is added to the
  plan, the \texttt{RE} ensures that requirements for the existence of
  the conditions and effects of that action (that is, the
  corresponding \texttt{LightBulb} must be off before and on after the
  action) are posted through the \texttt{PDB} and \texttt{CRE}
  modules.  }

@@

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different domains and problems, simply by providing
different models and goals. Consequently the expressiveness of the
language for models, goals and plans is of great importance. For \eue,
\nd allows the user to specify the different components of a problem
domain in a precise and concise manner. Among the main features that
make \nd a powerful tool for describing problem domains are that it is
object oriented. It supports classes and inheritance in similar
fashion to popular object-oriented languages like Java and C++. It
also supports polymorphism for some planning components. Using object
classes and instances is a time-tested approach to naturally describe
problem domains \cite{rumbaugh}. It offers declarative constructs to
define constraints and causality dependencies between actions and
effects and it offers procedural constructs to populate \eus
\texttt{PDB} with specific problem instances expressed in terms of
objects, variables, constraints, facts and goals.

To articulate the most important elements of \nde, we show how the
Shopping Agent problem could be modeled:

@@

\item[\textbf{Object Types}:] \nd supports an object-oriented
  approach, where common data and behavior can be abstracted as a
  class. In this case, we have the \textit{Product} class represent
  the products that the Shopping Agent needs, \textit{Location}
  represents the places where products can be found,
  \textit{ProductLocation} represents the association arising when a
  product can be purchased at a location. The Shopping Agent itself is
  represented by an \textit{Agent} class, which relies on the
  \textit{AgentLocation} class to keep track of where the agent is at
  any point in time. \nd supports a single rooted class hierarchy so
  that state and behavior can be extended through inheritance. We take
  advantage of that feature by making \textit{AgentLocation} inherit
  from \eus built-in \textit{Timeline} class which does not allow
  predicates to overlap in time for the same object instance. This
  feature is used for instance, to ensure that an agent can only be at
  one place at a time in our example.

@@

Domain rules can be defined in \nd in the context of a predicate or an
action. In the example above, the main elements of a rule definition
\kcomment{are as follows}:

@@

An \eu user can introduce new constraint types by implementing
specialized constraint and propagator classes. However \eu provides a
built-in constraint library with a range of arithmetic, temporal, set
and resource constraint types so that a large number of problem
domains can be modeled. Tables \ref{tab:calconst}, \ref{tab:setconst},
\ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
types included in \eus Constraint Library.  Temporal and resource
constraints are especially important to model real-world planning
domains.

\eu uses variables to explicitly represent timepoints for plan actions
and states. Constraints among timepoints provide a natural way to
express domain axioms. For example, in order to state that activity
$A$ must occur before activity $B$ we can say that the end timepoint
of $A$ is $leq$ the start timepoint of $B$.  \cite{dechter91} proposed
that constraints among timepoints can be grouped together to form a
Simple Temporal Network (\texttt{STN}). Such a network can be
transformed into a Distance Graph (\texttt{DG}) where the outward arc
from a node represents the maximum distance from the source node to
the target node.  \kcomment{They} also show that shortest path
algorithms \kcomment{(see \cite{corman} for a review)} could be used
to propagate values in the network and discover a negative cycle:
namely a path from a node to itself that has a path length less than
$0$. If such a cycle exists, the network is \kcomment{provably}
inconsistent. It was further shown that a single-source shortest path
algorithm was sufficient to detect a negative cycle and provide
sufficient propagation to yield a backtrack-free search. \eus
propagator for temporal constraints is an implementation of such a
single-source algorithm. Fig. \ref{fig:stn} illustrates an
\texttt{STN} with $2$ variables and a single constraint along with the
resulting \texttt{DG}.

\begin{figure}[!htb]
  \centering
  \subfloat[\small A Simple Temporal Network (\texttt{STN}). \kcomment{Nodes correspond
  to variables and their domains and arc's the \emph{constraint} or
  distance between them.}]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
  \caption{\small An illustration of a \texttt{STN} and its
    corresponding distance graph.}
  \label{fig:stn}
\end{figure}

Most real-world planning problems need to deal with plans made up of
actions that typically depend upon limited resources (\eg fuel,
storage capacity, battery state of charge, data buffers etc).  % \eu
% provides built-in support for resource computation.
In \eus representation, resource object types come in three forms with
related constraints. \textbf{Reservoir} resources can have both
\texttt{consume(resource,time,quantity)} and
\texttt{produce(resource,time,quantity)} constraints; production and
consumption is instantaneous. \textbf{Reusable} resources on the other
hand, have constraints that use a quantity of the resource for a
specific duration (\texttt{uses(resource,start,duration,quantity)}),
which are then subsequently released. \textbf {Unary} resources are
reusables with capacity$=1$, which \kcomment{accounts for} runtime
savings in comparison to the generic reusable class. Configuration
options are offered such that in addition to the usual constraint
violations when a resource is completely depleted, plan flaws can be
generated whenever a resource level is outside a specific bound, or
when instantaneous production or consumption exceeds a specific
threshold.

Determining whether a resource is the source of a plan flaw or a
constraint violation in general is substantially more difficult in
\eus flexible representation, than just computing a profile of the
resource and checking against some limits.  With temporal flexibility,
a number of resource profiles could be applicable depending on
precisely when each production/consumption event occurs out of a range
of possible values that are in the domain of the variables that
represent the temporal extent for the event.
\cite{Muscettola04,Muscettola06} propose an algorithm where the
variables that represent the time for production/consumption events
are placed in a temporal network, similar to the network used by the
temporal propagator; constraints on the resource can then be seen as
quantities \emph{flowing} through that network. Computing the maximum
flow then, yields the maximum resource usage and thus a lower bound on
the resource level; computing a minimum flow will similarly yield an
upper bound. They show that such upper and lower bound envelopes on
the resource level can be computed in polynomial time
\cite{corman}. Resource envelopes can be useful when trying to
evaluate plan brittleness in the face of changing activity durations
or start times. They can also be useful when searching for plans;
however they are relatively expensive to compute. Therefore their
ability to prune the search space must be weighted against the ability
to search faster \cite{6007772, morris11}. Fig. \ref{fig:resenvelopes}
shows an example of such a resource computation.

\begin{figure}
\centering
\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
\caption{\small An example of resource envelopes \cite{Muscettola04,Muscettola06} computed by \eue.}
\label{fig:resenvelopes}
\end{figure}

@@

Open conditions and threats allow flaw detection and resolution at a
higher-level of abstraction (\ie in terms of objects and tokens) than
that of binding variables as is common in Constraint Satisfaction
Problems. This is advantageous when a solver applies heuristics for
ordering choices since it provides a richer context in which to make
decisions.  Furthermore, it aids in reducing the amount of work done
by a solver so that only the necessary refinements are made,
\kcomment{ensuring} a partial plan with sufficient flexibility. For
example, one can bypass committing unbound timepoint variables since
threats will force a solver to impose ordering restrictions on these
variables.  Thus the planning process may yield partially-ordered
plans for which all possible extensions are provably valid.

The Plan Space Planning algorithm described can be implemented in
multiple ways depending on the approach chosen for flaw and resolver
selection and for backtracking. \eus built-in solver implements a
chronological backtracking algorithm that is summarized in Algorithm
\ref{alg:europa:solve}.

\begin{algorithm}
\KwIn{PartialPlan $plan$}
\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
\Begin{
  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
  \BlankLine
  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
  \BlankLine
  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \comment{\{\emph{potential branching point(s)}\}}\;\label{li:decisionpt}
  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
    \lElse{ $decision.undo()$ \comment{\{\emph{possible backtracking point}\}}\; } \label{li:backtrack}
   }
   \Return \texttt{false} \; \label{li:noplan}
 }
\caption{$\mathrm{bool} ~ solve(plan)$\kcomment{: To deliberate and
    generate partial plans}}
\label{alg:europa:solve}
\end{algorithm}

@@

\begin{algorithm}
  \KwIn{ConstraintEngine $ce$}
  \KwIn{int $maxIter$}
  \Begin{
    int $curIter \leftarrow 0$ \;
    \While{$\left( ce.getViolations() > 0 \right) ~ \& ~ \left( curIter <
        maxIter \right)$ }{
      Variable $queenToMove \leftarrow getQueenWithMaxViolation(ce)$
      \;
      SortedSet$<$int$> moves \leftarrow getMoves(ce, queenToMove,
      curPos)$ \;
      bool $moved \leftarrow $ \texttt{false} \;
      \ForEach{int $newPos$ of $moves$}{
        $moved \leftarrow tryMove(queenToMove, curPos, newPos)$\;
        \lIf{$moved$}{ \textbf{break} \; }
      }
      \lIf{$! moved$}{
        $forceMove(queenToMove, curPos, moves.first())$ \;
      }
      $checkSolution()$ \tcp*{Evaluate if a new best solution}
      $curIter$\texttt{++} \;
    }
  }
  \caption{$solveNQueens(ce, maxIter)$\kcomment{: An $N$-Queens solver}}
  \label{algo:n-queens}
\end{algorithm}
